Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-14 main.tex
fonet

\newcommand{\graph}[1]{\mathcal{#1}}
\newcommand{\vertices}[1][]{\ifthenelse{\isempty{#1}}{V}{V_{\graph{#1}}}}
\newcommand{\edges}[1][]{\ifthenelse{\isempty{#1}}{E}{E_{\graph{#1}}}}
\newcommand{\bgraph}[1]{\mathcal{\vec{#1}}}
\newcommand{\lgraph}[1]{\mathcal{L_{#1}}}
\newcommand{\leaps}[1]{L_{#1}}

@@

\newcommand{\lgG}{\lgraph{\gG}}

\newcommand{\lpG}{\leaps{\gG}}

@@

A graph $\gG$ is \bfit{labelled} in a set $L$ if each vertex $v\in\vG$
has an element $\ell(v)\in L$ associated with it, its \bfit{label}.  A
graph $\gG$ is (partially) \bfit{coloured} if it carries a partial
equivalence relation $\samecol$ on $\vG$; each equivalence class is a
\bfit{colour}.  A \bfit{vertex renaming} of $\gG=\tuple{\vG,\eG}$
along a bijection $(\hat\cdot)\colon\vG\to\hat\vG$ is the graph
$\hat\gG=\tuple{\hat\vG,\set{\hat v\hat w\mid vw\in\eG}}$, with
colouring and/or labelling inherited (i.e., $\hat v\samecol\hat w$ if
$v\samecol w$, and $\ell(\hat v)=\ell(v)$). Following standard graph
theory, we identify graphs modulo vertex renaming.

@@

Two atoms are \bfit{pre-dual} if their predicate symbols are dual
(e.g. $p(x, y)$ and $\dual{p}(y, z)$) and two literals are \bfit{pre-dual} if their
labels (atoms) are pre-dual. A \bfit{linked fograph} is a coloured fograph such
that every colour, called a \bfit{link}, consists of two pre-dual literals, and
every literal is either $\ttt$-labelled or in a link.

Let $\gG$ be a linked fograph. The set of links can be seen as a unification problem
by identifying dual predicate symbols. A \bfit{dualizer} of $\gG$ is an assignment
unifying all the links of $\gG$. Since a first-order unification problem is either
unsolvable or has a most general unifier, we can define the notion of \bfit{most
general dualizer}. A \bfit{dependency} is a pair $(\single x, \single y)$ of an
existential binder $\single x$ and a universal binder $\single y$ such that the most
general dualizer assigns to $x$ a term containing $y$. A \bfit{leap} is either a
link or a dependency. The \bfit{leap graph} $\lgG$ of $\gG$ is the undirected
graph $(\vG, \lpG)$ where $\lpG$ is the set of leaps of $\gG$. A set $W \subseteq \vG$ induces a \bfit{matching} in
$\gG$ if for all $w \in W$, $N(w) \cap W$ is a singleton, and induces a
\bfit{bimatching} in $\gG$ if it induces a matching in $\gG$ and in the leap graph
$\lgG$ as well.

\todo{define matching} \juihsuan{done in the cograph part}\lutz{no, it should be here}
\juihsuan{fixed}

\todo{adjust notation}

@@

\subsection{Skew Bifibrations}

A graph homomorphism $f: (\vG, \eG) \rightarrow (\vH, \eH)$ is a \bfit{fibration} if
for all $v \in \vG$ and $wf(v) \in \eH$, there exists a unique $\tilde{w} \in
\vG$ such that $\tilde{w}v \in \eG$ and $f(\tilde{w}) = w$, and is a \bfit{skew fibration}
if for all $v \in \vG$ and $wf(v) \in \eH$ there exists $\tilde{w} \in \vG$ such that
$\tilde{w}v \in \eG$ and $f(\tilde{w})w \notin \eH$. A \bfit{fograph homomorphism}
is a graph homomorphism between the underlying graphs of the fographs
considered. A fograph homomorphism $f : \gG \rightarrow \gH$ is
\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
\bfit{existential-preserving} if for all existential binder $b$ in $\gG$, the
vertex $f(b)$ is an existential binder in $\gH$.

\begin{definition}
A \bfit{skew bifibration} between fographs is a label-preserving and
existential-preserving fograph homomorphism whose restriction to the
binding graphs is a fibration.
\end{definition}

@@

\begin{definition}
A \bfit{first-order combinatorial proof} (\bfit{FOCP}) of a fograph $\gG$ is a skew
bifibration $f: \gN \rightarrow \gG$ where $\gN$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
\end{definition}

@@

Let $\Gamma$ be a sequent in $\FOMLL$. Consider the formula tree
$\formtree{\form{\Gamma}}$. A \bfit{link} on $\Gamma$ is a pair of
leaves whose atoms are pre-dual. A \bfit{linking} $\lambda$ on $\Gamma$ is a set
of
disjoint links such that each leaf of $\Gamma$ is either labelled by $\ttt$ or
in exactly one link. Similar to the set of links in linked fographs, a linking can be seen as a
unification problem, and a dualizer of the linking $\lambda$ is an assignment
unifying all the links in $\lambda$. There exists a \bfit{most general dualizer}
 of $\lambda$ if $\lambda$ has a dualizer. \juihsuan{Now I use the same terminology as for linked fographs}
A \bfit{dependency} is a pair $(\single \exists x, \single \forall y)$ of nodes
such that the most general dualizer assigns to $x$ a term containing $y$.

>>>>

\newcommand{\graph}[1]{\mathcal{#1}}
\newcommand{\vertices}[1][]{\ifthenelse{\isempty{#1}}{V}{V_{\graph{#1}}}}
\newcommand{\edges}[1][]{\ifthenelse{\isempty{#1}}{E}{E_{\graph{#1}}}}
\newcommand{\bgraph}[1]{\mathcal{\vec{#1}}}
\newcommand{\lgraph}[1]{\mathcal{#1}^{\mathsf{L}}}
\newcommand{\leaps}[1]{L_{#1}}

@@

\newcommand{\lgG}{\lgraph{\gG}}
\newcommand{\lgC}{\lgraph{\gC}}

\newcommand{\lpG}{\leaps{\gG}}
\newcommand{\lpC}{\leaps{\gC}}

@@

A graph $\gG$ is \bfit{labelled} in a set $L$ if each vertex $v\in\vG$
has an element $\ell(v)\in L$ associated with it, its \bfit{label}.  A
graph $\gG$ is (partially) \bfit{coloured} if it carries a partial
equivalence relation $\linkingof\gG$ on $\vG$; each equivalence class is a
\bfit{colour}.  A \bfit{vertex renaming} of $\gG=\tuple{\vG,\eG}$
along a bijection $(\hat\cdot)\colon\vG\to\hat\vG$ is the graph
$\hat\gG=\tuple{\hat\vG,\set{\hat v\hat w\mid vw\in\eG}}$, with
colouring and/or labelling inherited (i.e., $\hat v\samecol\hat w$ if
$v\samecol w$, and $\ell(\hat v)=\ell(v)$). Following standard graph
theory, we identify graphs modulo vertex renaming.

@@

Two atoms are \bfit{pre-dual} if their predicate symbols are dual
(e.g. $p(x, y)$ and $\dual{p}(y, z)$) and two literals are \bfit{pre-dual} if their
labels (atoms) are pre-dual. A \bfit{linked fograph} is a coloured fograph $\gC$ such
that every colour (i.e., equivalence class of~$\linkingof\gC$), called a \bfit{link}, consists of two pre-dual literals, and
every literal is either $\ttt$-labelled or in a link.

Let $\gC$ be a linked fograph. The set of links can be seen as a unification problem
by identifying dual predicate symbols. A \bfit{dualizer} of $\gC$ is a substitution
unifying all the links of $\gC$. Since a first-order unification problem is either
unsolvable or has a most general unifier, we can define the notion of \bfit{most
general dualizer}. A \bfit{dependency} is a pair $\set{\single x, \single y}$ of an
existential binder $\single x$ and a universal binder $\single y$ such that the most
general dualizer assigns to $x$ a term containing $y$. A \bfit{leap} is either a
link or a dependency. The \bfit{leap graph} $\lgC$ of $\gC$ is the undirected
graph $\tuple{\vC, \lpC}$ where $\lpC$ is the set of leaps of $\gC$. A vetex set $W \subseteq \vC$ induces a \bfit{matching} in
$\gC$ if for all $w \in W$, $N(w) \cap W$ is a singleton. We say that $W$ induces a
\bfit{bimatching} in $\gC$ if it induces a matching in $\gC$ and a matching in $\lgC$.

@@

\subsection{Skew Bifibrations}
\lutz{$\phi$ instead of $f$}
A graph homomorphism $f: \tuple{\vG, \eG} \rightarrow \tuple{\vH, \eH}$ is a \bfit{fibration} if
for all $v \in \vG$ and $wf(v) \in \eH$, there exists a unique $\tilde{w} \in
\vG$ such that $\tilde{w}v \in \eG$ and $f(\tilde{w}) = w$, and is a \bfit{skew fibration}
if for all $v \in \vG$ and $wf(v) \in \eH$ there exists $\tilde{w} \in \vG$ such that
$\tilde{w}v \in \eG$ and $f(\tilde{w})w \notin \eH$.
A directed graph homomorphism is a \bfit{fibration} if \todo{}

\lutz{do we really need that?
A \bfit{fograph homomorphism}
is a graph homomorphism between the underlying graphs of the fographs
considered. A fograph homomorphism $f : \gG \rightarrow \gH$ is
\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
\bfit{existential-preserving} if for all existential binder $b$ in $\gG$, the
vertex $f(b)$ is an existential binder in $\gH$.
}

\lutz{I think we need \bfit{fograph homomorphism} maps binders to binders and literals to literals, preserving the predicates.}

\begin{definition}
  Let $\gG$ and $\gH$ be forgraphs. A \bfit{skew bifibration} $\phi\colon\gG\to\gH$ is a fograph homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
  %etween fographs is a label-preserving and
%existential-preserving fograph homomorphism whose restriction to the
%binding graphs is a fibration.
\end{definition}

@@

\begin{definition}
A \bfit{first-order combinatorial proof} (\bfit{FOCP}) of a fograph $\gG$ is a skew
bifibration $\phi\colon \gC \rightarrow \gG$ where $\gC$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
\end{definition}

@@

Let $\Gamma$ be a sequent in $\FOMLL$. Consider the formula tree
$\formtree{\form{\Gamma}}$. A \bfit{link} on $\Gamma$ is a pair of
leaves whose atoms are pre-dual. A \bfit{linking} $\lambda$ on $\Gamma$ is a set
of
disjoint links such that each leaf of $\Gamma$ is either labelled by $\ttt$ or
in exactly one link. Similar to the set of links in linked fographs, a linking can be seen as a
unification problem, and a dualizer of the linking $\lambda$ is an assignment
unifying all the links in $\lambda$. There exists a \bfit{most general dualizer}
 of $\lambda$ if $\lambda$ has a dualizer. \juihsuan{Now I use the same terminology as for linked fographs}\lutz{use $\delta$ for the dualizer (or even better, make it a macro)}
A \bfit{dependency} is a pair $(\single \exists x, \single \forall y)$ of nodes
such that the most general dualizer assigns to $x$ a term containing $y$.
